---
title: "DonorHyperglycemiaStatsProject"
format: pdf
editor: source
---


## Running Code

```{r}
#| echo: false
#| error: false
#| warning: false
library (dplyr)
library(readr)
DonorHyperglycemia <- read_csv("DonorHyperglycemia.csv")
```

```{r}
#| echo: false
DonorHyperglycemia <- DonorHyperglycemia |>
  mutate(high_risk = if_else(donorrisk > 1.3, "yes", "no"))
```
## Age Hypothesis Test

```{r}
#| echo: false
t.test(r_age ~ delayed_fn, data = DonorHyperglycemia)
```

The null hypothesis is that the mean ages of both groups are equal, while the alternative hypothesis is that the mean ages are not equal. The test statistic follows a t distribution with 210.42 degrees of freedom. The p-value is 0.1065, which is above the significance level of 0.05. Therefore, we fail to reject the null hypothesis of the mean ages being the same, regardless of whether or not there was delayed liver graft function. 

## Sex Hypothesis Test:

The null hypothesis is that there is no relationship between sex and Delayed Liver Graft Function. The alternative hypothesis is that there is a relationship. 
```{r}
#| echo: false
chisq.test(table(DonorHyperglycemia$r_genderf, DonorHyperglycemia$delayed_fn))
```
The test statistic 0.41952 follows a Chi-squared distribution with one degree of freedom. The p-value of 0.5172 is greater than the significance level of 0.5, so we fail to reject the null hypothesis that there is no relationship. There is insufficient evidence to support there being a relationship between gender and delayed liver graft function.


